---
output:
  github_document:
    html_preview: false
---

<!-- README.md and index.md are generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

pkgload::load_all()

set.seed(20230702)

clean_output <- function(x, options) {
  x <- gsub("0x[0-9a-f]+", "0xdeadbeef", x)
  x <- gsub("dataframe_[0-9]*_[0-9]*", "      dataframe_42_42      ", x)
  x <- gsub("[0-9]*\\.___row_number ASC", "42.___row_number ASC", x)

  index <- x
  index <- gsub("─", "-", index)
  index <- strsplit(paste(index, collapse = "\n"), "\n---\n")[[1]][[2]]
  writeLines(index, "index.md")

  x <- fansi::strip_sgr(x)
  x
}

options(
  cli.num_colors = 256,
  cli.width = 80,
  width = 80,
  pillar.bold = TRUE
)

local({
  hook_source <- knitr::knit_hooks$get("document")
  knitr::knit_hooks$set(document = clean_output)
})
```


# pillar

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![R build status](https://github.com/r-lib/pillar/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-lib/pillar/actions)
[![Coverage status](https://codecov.io/gh/r-lib/pillar/branch/main/graph/badge.svg)](https://app.codecov.io/gh/r-lib/pillar)
[![CRAN status](https://www.r-pkg.org/badges/version/pillar)](https://cran.r-project.org/package=pillar)
<!-- badges: end -->


pillar provides tools for styling columns of data, artfully using colour and unicode characters to guide the eye.


## Installation

```{r, eval = FALSE}
# pillar is installed if you install the tidyverse package:
install.packages("tidyverse")

# Alternatively, install just pillar:
install.packages("pillar")
```


## Usage

pillar is a developer-facing package that is not designed for end-users.
It powers the `print()` and `format()` methods for [tibble](https://tibble.tidyverse.org/)s.
It also and defines generics and helpers that are useful for package authors who create custom vector classes (see https://github.com/krlmlr/awesome-vctrs#readme for examples) or custom table classes (like [dbplyr](https://dbplyr.tidyverse.org/) or [sf](https://r-spatial.github.io/sf/)).

```{r}
library(pillar)

x <- 123456789 * (10^c(-3, -5, NA, -8, -10))
pillar(x)

tbl_format_setup(tibble::tibble(x))
```

## Custom vector classes

The primary user of this package is [tibble](https://github.com/tidyverse/tibble), which lets pillar do all the formatting work.
Packages that implement a data type to be used in a tibble column can customize the display by implementing a `pillar_shaft()` method.

```{r}
library(pillar)

percent <- vctrs::new_vctr(9:11 * 0.01, class = "percent")

pillar_shaft.percent <- function(x, ...) {
  fmt <- format(vctrs::vec_data(x) * 100)
  new_pillar_shaft_simple(paste0(fmt, " ", style_subtle("%")), align = "right")
}

pillar(percent)
```

See `vignette("pillar", package = "vctrs")` for details.


## Custom table classes

pillar provides various extension points for customizing how a tibble-like class is printed.

```{r}
tbl <- vctrs::new_data_frame(list(a = 1:3), class = c("my_tbl", "tbl"))

tbl_sum.my_tbl <- function(x, ...) {
  c("Hello" = "world!")
}

tbl
```

See `vignette("extending", package = "pillar")` for a walkthrough of the options.

---

## Code of Conduct

Please note that the pillar project is released with a [Contributor Code of Conduct](https://pillar.r-lib.org/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
