---
title: "Customization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(pillar)
```

This draft describes a new backward-compatible design for making the appearance of a pillar more extensible.

The design consists of two parts:

1. Moving all formatting code into pillar
1. Adding extension points for formatting individual pillars


## Moving all formatting code into pillar

### Advantages

- Code that implements formatting for tables no longer needs tibble
- All formatting code in one package
- Simplify and document formatting code along the way
- New `as_tbl()` for attaching the `"tbl"` class to arbitrary objects (e.g. data frames or data tables) for printing

### Implementation

`format.tbl()` retains its semantics.
For compatibility it returns a character vector with one element per line.

```{r}
#' @export
format.tbl <- function(x, ...) {
  setup <- tbl_format_setup(x, width = width, n = n, n_extra = n_extra)
  header <- tbl_format_header(x, setup)
  body <- tbl_format_body(x, setup)
  footer <- tbl_format_footer(x, setup)
  c(header, body, footer)
}
```

### New generics

`tbl_format_setup()` takes the role of the (then superseded) `trunc_mat()`.
The default mimics `trunc_mat()` but returns a stable data structure that is well documented.
Classes that need to do more when setting up the printing implement a method.

Contrary to the existing logic, the width would be baked into the `setup` object.
I haven't seen a use case for formatting the same prepared object with multiple widths.

```{r}
tbl_format_setup.default <- function(x, ...) {
  trunc_mat(x, ...)
}
```

`tbl_format_header()` in the default implementation formats the output of `tbl_sum()`.
The implementer is expected to apply `style_subtle()` in case the class wants to add color or other styling.

```{r}
tbl_format_header.default <- function(x, ...) {
  sum <- tbl_sum(x)
  style_subtle(
    paste0(names(sum), ": ", sum)
  )
}
```

`tbl_format_body()` formats the colonnade prepared in the setup.
The implementation is shown further below.

`tbl_format_footer()` is copied from tibble.
The implementer is expected to apply `style_subtle()` in case the class wants to add color or other styling.

```{r}
tbl_format_footer.default <- function(x, ...) {
  style_subtle(format_footer(x, ...))
}

format_footer <- function(x, ...) {
  # Copy implementation from tibble
}
```

### Data flow

Boxes are functions and methods.
Solid lines are function calls.
Dotted lines are inputs to function calls.

```{r echo = FALSE}
DiagrammeR::mermaid("format.mmd")
```


## Adding extension points for formatting individual pillars

### Constraints and design goals

- A pillar is rarely shown individually, most often it is part of a colonnade.
- A colonnade also is rarely shown individually, most often it is part of a larger structure like a tibble.
- All pillars in a colonnade are shown in a similar way, a single controller can decide the appearance of all pillars.
- Pillars in a colonnade are either all named or all unnamed.
- Pillars can adapt their width to the available space. Computation of minimum and maximum width can happen before formatting the actual data. For performance reasons, `colonnade()` will not construct pillar objects it doesn't need.
- A pillar has a header (=capital), body (=shaft), footer (=basis, currently not used). Design should follow `cnd_header()`, `cnd_body()` and `cnd_footer()`. These components can be of different height, will be top-aligned when printing.
- Pillars are always shown from left to right, no "holes" in the colonnade. If the first column consumes all available space, the remaining columns are not shown, even if they all would fit if the first column is omitted.
- Printing pillars should take time proportional to the number of characters printed, and be "fast enough".
- Customizing parts of the display (e.g. omit type, add more information) should be easy.
- Existing infrastructure should be supported.


### Packed data frames and matrices

FIXME NEXT:

- Pillars are always constructed unnamed, do name separately (as a list of character vectors)
- Define ctl_* generics
- Implement and use replacement for colonnade(), internal, using the new generics
- Test coverage per file
- Improve output:
    - Tibble-local options for precision
    - Classes for numeric and string, use {formattable}
    - Multi-stage output for packed data frames
        - Reuse names/indices from left-hand side pillar if available
    - Arrays!
    - Show number of rows if known
    - Show columns that are abbreviated in full
- Look for FIXMEs


- Subdivide into name (perhaps multipart, shown in multiple rows and perhaps spanning multiple columns), and rest (type + data)
- Implement multipart pillars
- Enhance width attribute of pillar to support a vector
- Support subsetting of pillars

- Construct pillar objects while there still is space


### New generic

`colonnade()` gains a `controller` argument.
The default controller mimics current behavior.

A new method, `new_pillars()`, dispatch on that controller.
Apart from that, it receives the object, the formatted name (which can be a vector in the case of packed data frames or matrices), and the remaining width (which can also be a vector in the case of multiple tiers, if `getOption("width") < getOption("tibble.width")`).
It returns a list of `pillar` objects.

```{r}
new_pillars <- function(controller, x, width, ..., name = NULL) {
  UseMethod("new_pillars")
}

#' @export
new_pillars.tbl <- function(controller, x, width, ..., title = NULL) {
  if (is.data.frame(x)) {
    ...
  } else if (is.matrix(x)) {
    ...
  } else if (is.array(x)) {
    ...
  } else {
    list(pillar(x, title, width))
  }
}
```


### Implementation sketch

- We start with the assumption that each column consumes at least one character, with the space between the columns. This gives an upper bound on the number of columns and allows pre-allocating memory.
- For each column we could possibly fit, we call the `ctl_pillar_capital()` method and query the minimum width, until maximum width is reached.
- For each column where the capital fits, we call the `ctl_pillar_shaft()` method and query the minimum width, until maximum width is reached.
- Pillars are distributed using the existing algorithm.
