---
title: "Printing a tibble: Control and data flow"
output: rmdformats::downcute
vignette: >
  %\VignetteIndexEntry{Printing a tibble: Control and data flow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(pillar)
```

What happens when a tibble is printed?
This vignette documents the control flow and the data flow, explains the design choices, and shows the default implementation for the `"tbl"` class.
It is mainly of interest for implementers of table subclasses.
Customizing the formatting of a vector class in a tibble is described in `vignette("pillar", package = "vctrs")`.


## Requirements

- Fit into pre-specified width, distributing across multiple tiers if necessary
- Optionally shrink and stretch individual columns
- Header, body and footer for the tibble

    - Avoid recomputation of information

- Custom components for the pillars in a tibble, top-aligned

    - The container, not the column vectors, determine the appearance

- Customization of the entire output and of the pillars
- Support for data frame columns (packed data frames) and matrix/array columns


## Initialization

A tibble is a list of columns of class `"tbl_df"` and `"tbl"`.
Printing is designed to work for non-data-frame table-like objects such as lazy tables.
The `print.tbl()` method calls `format()` for the object and prints the output.
The `format.tbl()` method determines the width and the number of rows, creates a setup object, and uses that object to format header, body and footer.

```{r}
tbl <- tibble::tibble(a = 1:3, b = tibble::tibble(c = 4:6, d = 7:9), e = 10:12)
print(tbl, width = 23)
str(tbl)
```


```{r}
pillar:::print.tbl
```


```{r}
pillar:::format.tbl
```

While it's possible to extend or override these methods for your `"tbl"` subclass, often overriding the more specialized methods shown below is sufficient.

## Setup

Most of the work for formatting actually happens in `tbl_format_setup()`.
The desired output width is baked into the setup object and must be available when calling.
Setup objects print like a tibble but with a clear separation of header, body, and footer.

```{r}
setup <- tbl_format_setup(tbl, width = 24)
setup
```

A setup object is required here to avoid computing information twice.
For instance, the dimensions shown in the header or the extra columns displayed in the footer are available only after the body has been computed.

The default implementation converts the input to a data frame via `as.data.frame(head(x))`, and returns an object constructed with `new_tbl_format_setup()`.
If you override this method, e.g. to incorporate additional information, you can add new items to the default setup object, but you should not overwrite existing items.

```{r}
pillar:::tbl_format_setup.tbl
```

At the core, the internal function `ctl_colonnade()` composes the body.
Its functionality and the customization points it offers are detailed in the "Colonnade" section below.


## Header, body, footer

The components of a tibble are formatted with `tbl_format_*()` functions.
These functions return a character vector, with one element per line printed.
The setup object is required.

```{r}
tbl_format_header(tbl, setup)
tbl_format_body(tbl, setup)
tbl_format_footer(tbl, setup)
```

(The body is returned as a classed object with a `print()` method, it is still a `character()` under the hood.)

```{r}
class(tbl_format_body(tbl, setup))
typeof(tbl_format_body(tbl, setup))
```

Since most of the work already has been carried out in `tbl_format_setup()`, the default implementations mostly consist of code that styles and wraps the output.

```{r}
pillar:::tbl_format_header.tbl
```

```{r}
pillar:::tbl_format_body.tbl
```

```{r}
pillar:::tbl_format_footer.tbl
```


## Colonnade

The internal function `ctl_colonnade()` composes the body.
It performs the following tasks:

1. Create a pillar object for every column that fits, using `ctl_new_compound_pillar()`, `ctl_new_pillar()` and ultimately `pillar()` and `pillar_shaft()`
1. Determine the number of tiers and the width for each tier
1. Distribute the pillars across the tiers, assigning a width to each pillar.
1. Format each pillar via its `format()` function, passing the now known width.
1. Combine the formatted pillars horizontally.
1. Combine the tiers vertically.
1. Return the formatted body, and the columns that could not fit.

In the following, the first and the fourth steps are discussed.

## Creating pillar objects

Each column in the tibble is passed to `ctl_new_compound_pillar()`.
This is a generic that dispatches over the container:

```{r}
ctl_new_compound_pillar(tbl, tbl$a, width = 20)
ctl_new_compound_pillar(tbl, tbl$b, width = 20)
```

The default method distinguishes between compound and simple pillars.
Data frame, matrix, and array columns are decomposed into sub-pillars and returned as a compound pillar.
Regular vectors are forwarded to `ctl_new_pillar()`.
Implementers of `"tbl"` subclasses will rarely if ever need to extend or override this method.

```{r}
pillar:::ctl_new_compound_pillar.tbl
```

The `ctl_new_pillar()` method also dispatches over the container.

```{r}
pillar:::ctl_new_pillar.tbl
```

The default method simply calls `pillar()`.
This function now returns `NULL` if the width is insufficient to contain the data,
otherwise (ultimately) a pillar object constructed with `new_pillar()`.
It is possible to change the appearance of pillars by overriding or extending `ctl_new_pillar()`.

```{r}
pillar
```

By default, pillars consist of title, type, and body (shaft).
Formatting for title and type is provided by `new_pillar_title()` and `new_pillar_type()`.
The body can be customized by implementing `pillar_shaft()` for a vector class, see `vignette("pillar", package = "vctrs")` for details.
If title or type don't fit the available width, `pillar_shaft()` is never called.

## Formatting pillars

- `format.pillar()` with known width
- careful, don't align twice
